\documentclass[sigconf,anonymous,review]{acmart}

\settopmatter{printacmref=true}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage{xspace}
\usepackage{enumitem}
\setlength{\emergencystretch}{1em}

\newcommand{\method}{CARGO-AnoT\xspace}
\newcommand{\anot}{AnoT\xspace}
\newcommand{\tkg}{temporal knowledge graph\xspace}
\newcommand{\tkgs}{temporal knowledge graphs\xspace}
\newcommand{\fbeta}{$F_{0.5}$\xspace}

\title[CARGO-AnoT]{An Agentic Conflict-Aware Rule Graph Framework for Interpretable Online Anomaly Detection in Temporal Knowledge Graphs}
\acmConference[WSDM '27]{The 20th ACM International Conference on Web Search and Data Mining}{February 15--19, 2027}{Hong Kong, China}
\acmYear{2027}

\begin{document}

\begin{abstract}
Temporal knowledge graphs receive facts whose triples may be plausible even when their timestamps, entity roles, or supporting context are not. AnoT detects such updates with an interpretable rule graph learned from historical facts, but a graph of normal regularities alone gives limited evidence for different error mechanisms. We present \method, an error-aware extension of rule-graph anomaly detection. The method retains AnoT's normal rule graph and adds three type-specific evidence channels: role-consistency evidence for conceptual errors, contrastive temporal-gap rules for time errors, and train-only support rules for missing errors. A typed candidate--evidence--validation--scoring workflow records provenance for every proposed rule, verifies its evidence deterministically, and produces explanations only from accepted artifacts. On ICEWS14s, \method improves the reproduced AnoT baseline from 0.806/0.790/0.910 to 0.851/0.829/0.969 for conceptual errors and from 0.672/0.672/0.788 to 0.779/0.775/0.866 for time errors in Precision/\fbeta/AUC. For missing errors, it improves Precision and AUC from 0.726/0.729 to 0.751/0.822, while \fbeta remains sensitive to calibration. The implementation records the input evidence, selected calibration path, and final score for each evaluated fact.
\end{abstract}

\begin{CCSXML}
<ccs2012>
  <concept>
    <concept_id>10002951.10003317.10003347.10003350</concept_id>
    <concept_desc>Information systems~Data mining</concept_desc>
    <concept_significance>500</concept_significance>
  </concept>
  <concept>
    <concept_id>10010147.10010257.10010293.10010294</concept_id>
    <concept_desc>Computing methodologies~Knowledge representation and reasoning</concept_desc>
    <concept_significance>500</concept_significance>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Information systems~Data mining}
\ccsdesc[500]{Computing methodologies~Knowledge representation and reasoning}
\keywords{temporal knowledge graph, anomaly detection, rule graph, explainable data mining, online data quality}

\maketitle

\section{Introduction}

Temporal knowledge graphs record facts as time-stamped quadruples $(s,r,o,t)$. They are useful for event streams, financial records, and other evolving collections in which an update affects later reasoning. The problem is that incoming facts do not fail in one uniform way. A fact may place an entity in an implausible relational role, preserve a plausible triple while moving it to an implausible time, or reveal that an expected fact is absent. A practical detector therefore needs more than a ranking score. It needs an explanation that identifies the historical evidence behind the score and a protocol that prevents unsupported rules from silently entering the detector.

AnoT addresses this setting with a detector--updater--monitor architecture that summarizes a \tkg into a rule graph and traverses that graph to score new facts \cite{Zhang_2024}. This is a useful alternative to opaque temporal embeddings: each prediction can be related to typed static rules and temporal rule edges. Yet the normal rule graph primarily answers whether a candidate resembles historical regularities. That question is insufficient when the candidate itself is locally plausible. A political statement, for example, may be a common event, while the same statement becomes suspicious because it is separated from its historical context by an atypical interval. Likewise, an unseen entity pair is weak evidence of a conceptual error if the entities have previously played compatible roles in the relation.

This paper extends the normal-rule perspective with error-aware evidence. We retain the original AnoT graph as the common backbone and learn a small, auditable evidence layer for each error type. For time errors, the layer is mined by comparing normal training contexts against temporally displaced training facts constructed under the original AnoT perturbation protocol. For conceptual errors, the layer measures whether an entity has historically appeared in the subject or object role of the observed relation. For missing errors, it retrieves train-only contextual patterns that support the expected target fact. These signals have different semantics and are therefore not forced into a single conflict definition.

We call the resulting framework \method (Conflict-Aware Rule Graph Orchestration for AnoT). The word ``agentic'' has a narrow meaning here. It refers to a typed, artifact-based workflow with explicit Candidate, Evidence, Validation, Scoring, Explanation, Evaluation, and Reflection components. It does not mean that a language model decides whether a fact is true. The current experiments use deterministic candidate sources and programmatic checkers; an optional LLM-candidate schema is not used to produce the reported metrics.

Our contributions are as follows.
\begin{itemize}[leftmargin=*,itemsep=2pt,topsep=2pt]
  \item We formulate an error-aware rule-graph interface that preserves AnoT's normal evidence while exposing distinct conflict and support channels for conceptual, time, and missing errors.
  \item We introduce contrastive temporal-gap mining for time errors. The miner creates training-only time-displaced references, selects atomic-rule gap patterns with MDL, enrichment, and Bayesian direction tests, and freezes the resulting rules before test evaluation.
  \item We implement an auditable rule-evidence workflow with typed artifacts, input/output hashes, deterministic validators, and evidence-grounded explanations. This workflow separates rule discovery from final scoring and makes the provenance of accepted rules inspectable.
\end{itemize}

\section{Background and problem setting}

Let $\mathcal{G}=\{(s,r,o,t)\}$ be a \tkg, where $s,o\in\mathcal{E}$, $r\in\mathcal{R}$, and $t$ is a discrete timestamp. In online anomaly detection, historical facts are used to build a detector, after which an arriving fact $x=(s,r,o,t)$ receives an anomaly score. A larger score indicates weaker compatibility with the learned graph. The rule-based formulation complements temporal knowledge graph forecasting methods that use temporal logical patterns for explainable prediction \cite{Liu_2022} and sits within the broader literature on temporal knowledge graph reasoning \cite{Cai_2023}.

We follow AnoT's three error categories \cite{Zhang_2024}. A conceptual error makes the entity--relation combination implausible. A time error preserves a plausible entity--relation combination but places it at an implausible timestamp. A missing error is a held-out true fact whose surrounding history should support its presence. The categories matter because their evidence directions differ: conceptual and time errors need evidence that raises suspicion, whereas a missing target should receive a higher score when contextual support is present.

AnoT maps a fact to typed static rules of the form $(\tau_s,r,\tau_o,d)$, where $\tau_s$ and $\tau_o$ are entity types and $d\in\{\mathrm{in},\mathrm{out}\}$ is an orientation. Temporal edges connect static rules that occur in historical sequence. The graph is learned with a minimum-description-length objective and supplies reachable rules as explanations during scoring \cite{Zhang_2024}. \method uses this graph unchanged as the normal-evidence backbone. This choice is deliberate: conventional rule mining seeks compact, high-support regularities \cite{Galarraga_2015}, whereas our additional layer asks a different question---which rule contexts distinguish a specific error mechanism from normal updates.

For an evaluation item $x$, let $y_k(x)\in\{0,1\}$ denote its label for error type $k\in\{\mathrm{conceptual},\mathrm{time},\mathrm{missing}\}$. The label direction follows the original benchmark: a corrupted conceptual or temporal update has label one, while a held-out true fact has label one for the missing-fact task. Thus, the same graph observation can have different semantics across tasks. A long gap is evidence against the plausibility of a time-stamped update, but contextual recurrence is evidence \emph{for} an omitted true fact. This distinction is why we retain three evidence channels rather than relabeling all graph signals as ``conflicts.''

\section{CARGO-AnoT}

\subsection{Evidence interface over a normal rule graph}

For a candidate fact $x$, AnoT produces a base score $a(x)$ and a set of projected and reachable rules. \method converts these rule traversals into a named evidence vector $\phi(x)$. The generic portion contains rule coverage, support, missing-precursor, temporal-gap, successor-conflict, relation-pair conflict, and subject-state conflict features. Each feature is computed by deterministic graph traversal in \texttt{rule\_graph\_features.py}. For example, the support feature is the fraction of expected predecessor rules that are observed; the missing-precursor feature is its complement. A temporal-gap feature compares the observed gap with historical gap statistics, and successor conflict measures disagreement between projected rules and their reachable successors.

For the explicit-fusion paths, the final score combines the AnoT score and active evidence features:
\begin{equation}
S_k(x)=\alpha_k a(x)+\sum_{j\in\mathcal{F}_k}w_{k,j}\,\psi_{k,j}(x),
\label{eq:fusion}
\end{equation}
where $k$ denotes the error type, $\mathcal{F}_k$ is its feature set, and $\psi_{k,j}$ applies the type-specific sign convention. In particular, support is negative evidence for ordinary anomaly scoring but positive evidence for a missing fact. The evaluator selects weights and a decision threshold on validation and freezes them before test evaluation. For the conceptual \texttt{fair\_select} and optional missing calibration paths, the same named features are instead passed to a validation-selected linear, logistic, or shallow gradient-boosted calibrator. Thus, Equation~\ref{eq:fusion} describes the explicit path rather than every implementation branch.

Table~\ref{tab:evidence-interface} makes the interface explicit. The base graph and generic features are shared infrastructure, not separate detectors. The type-specific features are attached only to the matching task. This restriction prevents a feature with the wrong direction from being imported merely because it improved a different anomaly category on validation.

\begin{table}[t]
\centering
\caption{Evidence interface used by the full configuration. ``Positive direction'' means that a larger feature value increases the score for the stated task after applying $\psi_{k,j}$.}
\label{tab:evidence-interface}
\small
\begin{tabular}{@{}p{0.23\linewidth}p{0.39\linewidth}p{0.22\linewidth}@{}}
\toprule
Task & Active evidence & Positive direction \\
\midrule
Conceptual & Pair novelty; subject--relation and object--relation novelty; entity--relation novelty; role support; neighborhood overlap & Role incompatibility / novelty \\
Time & Base temporal features; validated contrastive atomic-gap rules & Conflict / displaced-gap match \\
Missing & Context coverage; target support; pair and entity--relation history; temporal support & Support for the held-out target \\
\bottomrule
\end{tabular}
\end{table}

The explicit fusion implementation applies the sign before weighting. For instance, conceptual relation-role support receives a negative sign because repeated role evidence makes a conceptual corruption less likely. In contrast, missing support receives a positive sign because a well-supported but absent target is the positive class in that task. The evaluator serializes every component value and the final score for each example; calibration metadata records the selected model when a learned path is used.

\subsection{Conceptual evidence from relational roles}

Entity-pair novelty alone is unreliable in event data because valid new pairs appear frequently. The conceptual module therefore builds train-graph profiles for subject--relation and object--relation counts, relation-specific pair history, and entity-neighborhood overlap. Given $(s,r,o,t)$, it derives features such as whether $s$ has previously occurred as the subject of $r$, whether $o$ has occurred as the object of $r$, and the normalized support for these roles. The implementation also includes low-neighbor-overlap and pair-relation novelty features.

The module can mine type-relation incompatibility rules from training perturbations, but the reported conceptual configuration uses a validation-selected \texttt{novelty\_role} feature set. The \texttt{fair\_select} implementation compares linear, logistic, and shallow gradient-boosted candidates on a temporal holdout within validation; its gradient-boosted candidates use $80$ estimators, depth $1$--$3$, learning rates $0.02$, $0.05$, or $0.1$, and random state $42$. No test labels take part in feature selection or calibration.

The role profile is intentionally finer than pair novelty. A previously unseen pair $(s,o)$ is common in event streams and is not, by itself, a reliable error signal. The useful question is whether $s$ has occurred in the subject role of $r$, whether $o$ has occurred in its object role, and whether their local relation neighborhoods are compatible with the observed role. The selected \texttt{novelty\_role} set contains pair novelty, pair--relation novelty, generated-sample novelty, subject--relation novelty, object--relation novelty, entity--relation novelty, relation-role support, and low-neighbor-overlap features. The reported time configuration uses explicit score fusion; missing support can use either its explicit path or its separately validated optional calibration path.

\subsection{Contrastive temporal-gap rules}

Time errors require a different signal. We construct training references that keep a fact's subject, relation, and object but place it at a later target time. The generator takes facts from the earliest ten training timestamps, requires that the fact's latest training occurrence precede the target by at least $200$ time units, and excludes facts known in validation or test. At each target time, it samples up to $30\%$ as many displaced references as normal facts. With ICEWS14s, this yields $25{,}794$ normal training samples and $7{,}654$ displaced references.

For each sample, the miner extracts a pattern
\begin{equation}
q=(r_{\mathrm{pre}},r_{\mathrm{tar}},[\ell,u)),
\end{equation}
where $r_{\mathrm{pre}}$ and $r_{\mathrm{tar}}$ are oriented atomic rules on the same entity pair and $[\ell,u)$ is a gap bucket. The default buckets are $[0,30)$, $[30,90)$, $[90,180)$, $[180,365)$, and $[365,\infty)$. Repeated-rule patterns are allowed: an old occurrence of the same atomic rule can be informative when it reappears at an implausible time.

For each pattern, we count unique normal and displaced training samples, calculate a label-partition MDL gain, and require a positive anomaly direction. Candidates must satisfy at least $10$ anomaly hits, $20$ total hits, lift at least $1.25$, AUC at least $0.505$, posterior direction probability at least $0.90$, a positive $95\%$ credible lower bound on the rate difference, and positive MDL gain. Validation recomputes the statistics on normal facts and AnoT time errors; it does not create new patterns. The selected-rule weight combines normalized training MDL gain, validation enrichment, and posterior direction probability.

More precisely, for a candidate $q$, let $h^+_q,n^+_q$ be its anomaly hits and total anomaly samples, and let $h^-_q,n^-_q$ be the corresponding normal quantities. The validation selector uses
\begin{equation}
\operatorname{AUC}_q=\frac{1}{2}\left(\frac{h^+_q}{n^+_q}+1-\frac{h^-_q}{n^-_q}\right),
\quad
\operatorname{lift}_q=
\frac{h^+_q/(h^+_q+h^-_q)}{n^+_q/(n^+_q+n^-_q)}.
\label{eq:rule-statistics}
\end{equation}
It additionally draws from independent Beta--Binomial posteriors for the two hit rates and requires $\Pr(p_q^+>p_q^-)\geq0.90$ together with a positive lower endpoint of the $95\%$ credible interval for $p_q^+-p_q^-$. The final bounded weight is proportional to
\begin{equation}
\widetilde{\operatorname{MDL}}_q\,\max\{0,\log_2(\operatorname{lift}_q)\}\,
\Pr(p_q^+>p_q^-),
\label{eq:rule-weight}
\end{equation}
where $\widetilde{\operatorname{MDL}}_q$ is the training MDL gain normalized among eligible rules. Equations~\ref{eq:rule-statistics}--\ref{eq:rule-weight} explain why a frequent but nondiscriminative temporal pattern is discarded rather than becoming a conflict rule.

\subsection{Discovery, validation, and freezing}

The temporal rule file is constructed in three stages. In discovery, the miner creates normal/displaced training batches and extracts all observed atomic-gap patterns. It first removes patterns that fail minimum hit counts, reverse the anomaly direction, have weak lift or AUC, or do not reduce the label-partition description length after paying the rule-encoding cost. In validation, it replays the \emph{fixed} candidate list on validation normal and time-error samples. This second stage computes the posterior tests in Equation~\ref{eq:rule-statistics}, ranks surviving rules by posterior direction, credible rate difference, AUC, and anomaly hits, and retains at most $500$ rules. Finally, it writes a serialized rule file. Test evaluation loads only this file; it never enumerates a new temporal pattern.

The distinction between discovery and validation is more than an implementation detail. A time-displaced reference is a controlled contrast source, not a claim that every displaced event is a real anomaly. The train stage uses it to propose hypotheses about typed contexts. The validation stage asks whether the same hypotheses separate the benchmark's independently generated time errors from normal facts. A candidate can be statistically prominent in the train reference data and still be rejected on validation. Conversely, a rule's eventual test contribution is traceable to a bounded candidate set selected before test scoring.

This procedure selected $500$ temporal-gap rules from $1{,}255$ candidates. Their union matched $65.4\%$ of validation time errors and $3.2\%$ of validation normal samples, with union AUC $0.811$. One representative pattern links an out-oriented \texttt{Make\_statement} atomic rule to the same rule with a $[180,365)$ gap. It does not assert that repeated statements are anomalous. It records that this typed, oriented, long-gap recurrence is much more frequent in the displaced reference context than in normal data.

\subsection{Train-only support for missing facts}

Missing-error evidence has the opposite polarity. The missing-support miner uses only training facts and represents a rule as $(r_{\mathrm{ctx}},r_{\mathrm{tar}},\rho,[\ell,u))$, where $r_{\mathrm{ctx}}$ is a contextual atomic rule, $r_{\mathrm{tar}}$ is a target rule, $\rho$ specifies a shared subject, object, or pair role, and $[\ell,u)$ is a gap bucket. It retains patterns with sufficient context and target support, then exposes contextual coverage, temporal support, pair history, entity-relation support, and broad support features. The primary missing configuration uses this train-only support graph; the optional contrastive missing-support variant is disabled because its validation behavior was not stable.

\subsection{Auditable candidate workflow and explanations}

The evidence workflow wraps the detector without replacing its decision rule. Candidate artifacts include source, scope, polarity, binding, gap interval, dataset hash, and provenance. CandidateAgent can package symbolic, contrastive, conceptual, missing-support, or optional LLM-proposed candidates. The reported time-rule audit uses deterministic candidate sources. EvidenceAgent retrieves matched historical facts and rule evidence. ValidationAgent accepts a candidate only when its validation evidence passes the configured enrichment, AUC, Beta--Binomial direction, credible-difference, and MDL checks. ScoringAgent converts accepted candidates into the rule JSON consumed by the CARGO-AnoT evaluation path.

Every agent inherits a common trace writer that records input and output hashes, tool calls, validation status, runtime, and failure reason. ExplanationAgent produces a natural-language explanation only from accepted evidence artifacts. ExplanationVerifierAgent then checks that the target fact, historical fact, rule identifier, and time gap in the explanation are all present in the evidence record. The workflow therefore treats an explanation as a checked view of a score trace, not as a separate generator of reasons.

The workflow has eight fixed responsibilities. CandidateAgent emits a versioned candidate specification; EvidenceAgent retrieves historical matches; ValidationAgent applies split-specific tests; ScoringAgent writes the frozen rule artifact; ExplanationAgent renders an accepted artifact for a user; ExplanationVerifierAgent checks every referenced field; EvaluationAgent records scores and metrics; and ReflectionAgent aggregates failure reasons for subsequent development. Only the first four stages can change the rule file, and only before test evaluation. The later stages may inspect artifacts but cannot add a rule or update the graph. This separation is important in an online setting, where an explanation or a post-hoc diagnosis must not create an unlogged feedback path into the detector.

The workflow is deliberately typed rather than prompt-driven. A candidate specifies its source, error polarity, rule binding, temporal interval, dataset scope, and provenance. An evidence record lists the candidate identifier, target fact, matched historical fact, observed interval, and label/split context. A validation record holds the computed statistics and explicit rejection reasons. The scoring artifact contains only accepted rules and their bounded weights. This contract lets the code reject malformed or split-inconsistent artifacts before they influence a score. It also leaves room for a future LLM candidate proposer without making an LLM part of the present detection claim: any such proposal would still need the same evidence retrieval and deterministic validation.

\section{Experimental protocol}

\subsection{Data, task construction, and protocol boundary}

We evaluate ICEWS14s with the original AnoT train/validation/test split files and anomaly-generation protocol. The source data are represented as integer quadruples and grouped by discrete timestamps. For each evaluation timestamp, normal facts are the observed split facts. Conceptual negatives are generated by replacing a compatible entity or relation with one that is not supported by the training graph; temporal negatives retain $(s,r,o)$ but place an early training fact at a later timestamp; and missing positives are sampled held-out facts, while their negatives combine time and conceptual corruptions. The generator seeds Python and NumPy with $42$, uses conceptual and missing ratios of $0.6$, a time ratio of $0.3$, and the ICEWS14s temporal span of $200$. Its temporal candidate pool is restricted to the earliest ten training timestamps and excludes facts known in validation or test, matching the original early-candidate construction.

This protocol produces labels for evaluation; it does not make the test labels available to the proposed miner. The contrastive time module uses the same construction \emph{inside training} to define a reference distribution for rule discovery. It does not use validation or test anomaly labels during training. Validation contains the only labels used for candidate selection, feature selection, score weights, and thresholds. Test scoring then consumes frozen rule files with graph updates disabled. The code-level guard rejects a training path that reads validation/test artifacts except as a known-fact collision filter, rejects candidate selection outside the validation split, and rejects test evaluation with updates or an unfiltered candidate file.

\begin{table}[t]
\centering
\caption{Key implementation settings for the reported ICEWS14s configuration. All temporal-rule settings are specified in \texttt{contrastive\_conflict\_mining.py} and \texttt{contrastive\_conflict\_selection.py}.}
\label{tab:implementation}
\small
\begin{tabular}{@{}p{0.55\linewidth}p{0.27\linewidth}@{}}
\toprule
Setting & Value \\
\midrule
Random seed & 42 \\
Original AnoT temporal span / early timestamps & 200 / 10 \\
Training displaced-reference ratio & 0.30 \\
Gap boundaries & 0, 30, 90, 180, 365 \\
Training minimum anomaly / total hits & 10 / 20 \\
Validation minimum lift / AUC & 1.25 / 0.505 \\
Posterior direction / credibility & 0.90 / 0.95 \\
Maximum temporal candidates / selected rules & 2,000 / 500 \\
Conceptual calibration candidates & linear, logistic, or GB (80 trees; depth 1--3; lr 0.02/0.05/0.1) \\
Manual conflicts / missing contrastive support & disabled / disabled \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Baselines, calibration, and measures}

We compare \textbf{AnoT}, the reproduced original detector with validation-tuned thresholds, against \textbf{CARGO-AnoT}, the full error-aware configuration. CARGO-AnoT uses contrastive temporal rules for time, \texttt{novelty\_role} calibration for conceptual errors, and train-only missing support for missing errors. Manual relation-conflict rules are disabled because they did not survive validation in the time workflow. The optional missing-contrastive-support module is also disabled in the reported configuration because its validation behavior was unstable.

For explicit fusion, validation searches the configured feature-weight grid and selects the threshold that maximizes \fbeta. The conceptual \texttt{fair\_select} path avoids choosing a feature set on test data: it evaluates candidate linear and shallow gradient-boosted calibrators on a temporal holdout inside validation, selects one specification, then refits that specification on validation before scoring test. We report Precision, \fbeta, and ROC-AUC. Precision and \fbeta depend on the selected threshold, whereas AUC measures ranking independently of a single operating point. This distinction is central to interpreting the missing-fact results.

\subsection{Reproducibility and audit artifacts}

Each publication-mode evaluation emits the frozen configuration, source snapshot, selected-rule-file hashes, prediction rows, and a result-registry record. The registry also stores the protocol-guard outcomes, the command line, and the final metrics. Candidate, evidence, and validation artifacts carry their own input/output hashes. These artifacts do not replace an independent reproduction, but they make it possible to establish exactly which rule file and which settings produced a reported score. We use the archived registry values in this draft; before submission, the full table must be regenerated from one frozen archive and accompanied by the corresponding prediction files.

We report Precision, \fbeta, and AUC. \fbeta is the metric implemented by the evaluation code, with $\beta=0.5$. The full \method configuration uses contrastive temporal rules, conceptual role evidence, and train-only missing-support evidence. It disables manual relation-conflict rules and the unstable missing-contrastive-support module.

\section{Results}

Table~\ref{tab:main-results} gives the current recorded ICEWS14s results. The strongest gains occur where the evidence matches the error mechanism. Conceptual role evidence improves all three conceptual metrics. Contrastive temporal-gap evidence gives the largest relative improvement for time errors. The complete configuration improves missing-error Precision and AUC over AnoT, but its \fbeta is marginally lower. This is consistent with a calibration trade-off: the missing-support features improve ranking and high-score precision, while the current threshold loses recall.

\begin{table*}[t]
\centering
\caption{ICEWS14s test results. Each triplet is Precision / \fbeta / AUC. Bold indicates the better value in a column. CARGO-AnoT denotes the full error-aware configuration executed through the CARGO artifact workflow.}
\label{tab:main-results}
\small
\begin{tabular}{lccc}
\toprule
Model & Conceptual errors & Time errors & Missing errors \\
\midrule
AnoT & 0.806 / 0.790 / 0.910 & 0.672 / 0.672 / 0.788 & 0.726 / \textbf{0.658} / 0.729 \\
CARGO-AnoT & \textbf{0.851} / \textbf{0.829} / \textbf{0.969} & \textbf{0.779} / \textbf{0.775} / \textbf{0.866} & \textbf{0.751} / 0.657 / \textbf{0.822} \\
\bottomrule
\end{tabular}
\end{table*}

The temporal miner itself has strong validation separation. Table~\ref{tab:time-mining} distinguishes discovery from final testing. Training proposes patterns from displaced references, while validation determines whether they remain informative under the evaluation protocol. This separation matters because a rule that merely reproduces the artificial perturbation would not necessarily discriminate the validation time errors.

\begin{table}[t]
\centering
\caption{Temporal-gap mining audit on ICEWS14s. The candidate and selected-rule counts are rule artifacts; union statistics are computed on validation.}
\label{tab:time-mining}
\small
\begin{tabular}{lr}
\toprule
Quantity & Value \\
\midrule
Normal training samples & 25,794 \\
Displaced training references & 7,654 \\
Observed patterns & 170,444 \\
Candidates after training filters & 1,255 \\
Selected temporal-gap rules & 500 \\
Validation anomaly hit rate & 0.654 \\
Validation normal hit rate & 0.032 \\
Validation union AUC & 0.811 \\
\bottomrule
\end{tabular}
\end{table}

The candidate workflow also supplies a useful audit of what actually drives the final time model. Table~\ref{tab:workflow-audit} shows that none of the symbolic relation-pair or atomic-rule proposals survived validation under the configured filters. All accepted rules are contrastive atomic-gap rules. This finding is important for interpretation: the gain should be attributed to temporal displacement mining, not to a broad claim that every conflict source helps.

\begin{table}[t]
\centering
\caption{Validation audit of the time-error candidate workflow. No LLM-proposed candidates were used in the reported run.}
\label{tab:workflow-audit}
\small
\begin{tabular}{lrr}
\toprule
Candidate source and scope & Proposed & Accepted \\
\midrule
Symbolic relation pair & 500 & 0 \\
Symbolic atomic rule & 1,000 & 0 \\
Contrastive atomic gap & 500 & 500 \\
\midrule
Total & 2,000 & 500 \\
\bottomrule
\end{tabular}
\end{table}

\section{Analysis and discussion}

\subsection{Evidence quality follows the error mechanism}

The three channels should not be read as interchangeable feature engineering. They resolve three different ambiguities that remain after normal rule-graph scoring. First, for conceptual errors, pair novelty is too coarse: new entity pairs occur in normal political event streams. The role features ask whether an entity has ever occupied the relevant subject or object role. In the recorded ICEWS14s analysis, the object--relation-unseen indicator is active for $86.1\%$ of conceptual positives but only $17.1\%$ of normal references. This separation is consistent with the simultaneous improvement in Precision, \fbeta, and AUC under the selected role-aware calibration.

Second, a raw temporal gap is not necessarily anomalous. Some relations recur after long intervals, while others do not. The contrastive miner therefore conditions on the predecessor and target atomic rules as well as the interval bucket. Its selected-rule union covers $65.4\%$ of validation time errors and only $3.2\%$ of normal references. The resulting union AUC is $0.811$, before this signal is combined with the AnoT score. The contrastive feature is consequently evidence about a particular context--gap pattern, not a global rule that ``long intervals are anomalous.''

Third, the missing task reverses the usual anomaly semantics. A score should rise when history supports an expected target that is absent, not when history conflicts with a supplied fact. The train-only support graph improves AUC from $0.729$ to $0.822$ and Precision from $0.726$ to $0.751$ relative to AnoT, but the full configuration obtains $0.657$ \fbeta versus $0.658$ for AnoT. The result is useful but qualified: the channel improves ranking and high-confidence selection, whereas the current threshold loses enough recall to erase the \fbeta gain.

\begin{table}[t]
\centering
\caption{Diagnostic view of the current ICEWS14s results. Deltas compare CARGO-AnoT against AnoT; values are derived directly from Table~\ref{tab:main-results}.}
\label{tab:diagnostics}
\small
\begin{tabular}{lrrr}
\toprule
Error type & $\Delta$ Precision & $\Delta$ \fbeta & $\Delta$ AUC \\
\midrule
Conceptual & +0.045 & +0.039 & +0.059 \\
Time & +0.107 & +0.103 & +0.078 \\
Missing & +0.025 & -0.001 & +0.093 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{What the agentic layer does and does not claim}

The word ``agentic'' can be misleading if it is used as a synonym for an LLM classifier. It is not one here. The reported candidates are deterministic symbolic, contrastive, conceptual, and missing-support artifacts. The agentic contribution is operational: it makes the transition from candidate to evidence, accepted rule, score contribution, and explanation explicit and checkable. Table~\ref{tab:workflow-audit} provides a concrete example. It shows that the implementation did not retain symbolic candidates simply because they were easy to enumerate. They were proposed, evaluated, and rejected; the accepted time rule file contains only contrastive atomic-gap rules.

This separation also limits the claim. The workflow verifies that an explanation's stated target, historical fact, rule identifier, and gap occur in the stored evidence record. It does not yet establish a user-study measure of explanation usefulness, nor does it show that workflow orchestration alone improves detection metrics. Those are separate empirical questions. Treating provenance and explanation checking as engineering guarantees rather than as unmeasured accuracy gains keeps the paper's claims aligned with the implementation.

\subsection{Failure modes and practical use}

The diagnostic table identifies two practical cautions. A time rule mined from a benchmark perturbation can overfit the perturbation mechanism if its scope is not tested on a fresh split or another dataset. The train--validation--test guards prevent direct label leakage, but they do not erase the risk created by repeated test-set feedback during research iteration. We therefore treat the current ICEWS14s results as the recorded development result and require a frozen rerun on fresh data before archival submission.

The missing-fact result highlights a second caution. AUC, Precision, and \fbeta answer different questions. The current support channel ranks true missing facts above negatives more reliably, but the validation-selected operating threshold is not yet recall-optimal. In a data-curation deployment, the right threshold depends on whether the operator is reviewing a small, high-confidence queue or trying to recover as many omissions as possible. The system should expose this choice rather than presenting a single threshold as universally optimal.

\section{Related work}

Graph anomaly detection covers anomalous nodes, edges, and subgraphs under structural, statistical, and learning-based formulations \cite{Akoglu_2014}. Temporal knowledge graph research has largely emphasized completion and forecasting, including logical-rule methods that provide paths or rules as explanations \cite{Liu_2022,Cai_2023}. These methods ask which future facts are likely to occur. Online anomaly detection has a different target: it must flag facts that should not be trusted while retaining an explanation useful for graph maintenance. AnoT is the closest base method because it learns a compact rule graph for online, interpretable anomaly scoring \cite{Zhang_2024}. \method keeps this rule-graph formulation rather than replacing it with a latent representation.

Rule miners such as AMIE+ search for compact, high-confidence symbolic implications in incomplete knowledge bases \cite{Galarraga_2015}. Our miner also returns a finite symbolic rule set, but its selection target is different: a temporal pattern must distinguish normal contexts from time-displaced references, then survive validation against evaluation-style time errors. Our temporal component is therefore not a generic contrastive representation learner. It uses contrast only to mine typed atomic-gap patterns and validates each candidate with explicit frequency, enrichment, posterior-direction, and MDL criteria. The result remains a rule file consumed by the same graph-based scorer. The workflow component is similarly conservative: its agents package, retrieve, verify, and explain artifacts. It does not delegate anomaly judgment to an unverified generative model.

The distinction from graph anomaly detection is also practical. A generic graph anomaly method often treats rarity, reconstruction error, or neighborhood inconsistency as the signal of interest \cite{Akoglu_2014}. In a temporal knowledge graph, an event can be rare but correct, or frequent but misplaced in time. CARGO-AnoT therefore does not use rarity as a sufficient explanation. Its score traces return a typed relation-role condition, a predecessor--target gap condition, or a context--target support condition. These conditions are narrower than a claim that an entire entity, relation, or subgraph is anomalous, which is appropriate for a curator deciding whether to inspect a single arriving quadruple.

Temporal link forecasting provides an adjacent but different form of temporal reasoning. Rule-based forecasters such as TLogic encode relation sequences that make a future link plausible \cite{Liu_2022}; survey work places these methods alongside embedding and neural approaches for temporal knowledge graph completion \cite{Cai_2023}. CARGO-AnoT reuses the useful part of this perspective---explicit temporal patterns---but changes the decision problem. A temporal-gap rule is not used to predict the most likely future object. It is used to assess whether a specified event occurs in a context and interval that are overrepresented among time-displaced references. The output remains an anomaly score with evidence, not a ranking over missing links.

Finally, the workflow differs from a free-form agent architecture. The term agentic describes ownership and traceability of intermediate artifacts, not autonomous problem reformulation. Candidate discovery is separable from evidence retrieval; evidence retrieval is separable from validation; and validation is separable from final scoring. This makes rejected rules informative rather than invisible. In the recorded time workflow, symbolic relation-pair and atomic-rule candidates were retained in the audit report even though none entered the final score. The distinction matters for research reporting: the paper can attribute the observed time gain to the validated contrastive atomic-gap source rather than suggesting that all conflict heuristics contributed.

\section{Limitations}

The reported evidence is concentrated on a single ICEWS14s protocol. The temporal displacement references intentionally resemble the benchmark's time-error construction, so the results support performance under this protocol rather than universal detection of real-world temporal errors. The current missing-error configuration improves ranking and precision but not \fbeta, which indicates that recall-sensitive calibration needs further work. The candidate workflow verifies evidence provenance and explanation faithfulness, but its operational benefit still needs evaluation on longer online streams with distribution shift. Finally, test results were inspected during development. Before a final archival submission, the method and hyperparameters should be frozen and evaluated on a fresh split or an additional dataset not used for iterative design decisions.

\section{Ethical considerations}

This work uses event data and evaluates synthetic anomaly types derived from an established benchmark protocol. The detector should be treated as decision support for data curators, not as an authority on political or social truth. Rule-based explanations can make a system easier to audit, but they can also expose historical biases in the source graph. A deployment should therefore retain human review for rejected or quarantined facts, log the evidence trace, and avoid using anomaly scores to make consequential decisions about people or organizations without independent verification.

\section{Conclusion}

\method extends AnoT with error-aware evidence while preserving its interpretable rule-graph backbone. Role consistency addresses conceptual errors, contrastive temporal-gap rules address time errors, and train-only support rules address missing facts. The typed workflow records how a rule was proposed, checked, and used in a score or explanation. On ICEWS14s, the method produces substantial gains for conceptual and time errors and clarifies the remaining calibration problem for missing errors. The next empirical step is a frozen evaluation on fresh data, together with ablations that measure the marginal value of each evidence channel and the operational value of the audit workflow.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
